### 2017-01-17 text/Chapter03.tex
Edits pages 31-94

\paragraph{Chapter structure and contributions}
\begin{itemize}
\item We characterize contextual properties -- the Section~\ref{sec:applications-structure}
  explains what is a \emph{coeffect} and contrasts it with a better known notion of
  \emph{effect}. It explains what is the nature of properties that can be tracked using
  coeffect systems presented in this thesis.

@@

\subsection{Effectful lambda abstraction}
\label{sec:applications-structure-lam}

The difference between effects and coeffects becomes apparent when we consider lambda abstraction.
The typical lambda abstraction rule for effect systems looks as (\emph{effect}) in
Figure~\ref{fig:applications-abs}. Wadler and Thiemann~\cite{monads-effects-marriage} explain how
the effect analysis works as follows:
%
\begin{quote}
\emph{In the rule for abstraction, the effect is empty because evaluation immediately
returns the function, with no side effects. The effect on the function arrow
is the same as the effect for the function body, because applying the function will
have the same side effects as evaluating the~body.}
\end{quote}
%
This is the key property of \emph{output impurity}. The effects are only produced when the
function is evaluated and so the effects of the body are attached to the function. A recent
work by Tate~\cite{effects-producer-semantics} uses the term \emph{producer} effect systems
for such standard systems and characterises them as follows:
%
\begin{quote}
\emph{Indeed, we will define an effect as a producer effect if all computations with that
effect can be thunked as ``pure'' computations for a domain-specific notion of purity.}
\end{quote}
%
The thunking is typically performed by a lambda abstraction -- given an effectful expression
$e$, the function $\lambda x.e$ is an effect free value (thunk) that delays all effects.
As shown in the next section, contextual properties do not follow this pattern.

@@

The interesting case is when we construct a lambda function $\lambda x.e$, marshall it and
send it to another node. In systems such as Acute \cite{app-distributed-acute}, the context
requirements can be satisfied in a number of ways. When the same resource is available at the target
machine (\eg~current time), we can transfer the function with a context requirement and \emph{rebind}
the resource. However, if the resource is not available (\eg.~GPS on the server), we need to a
capture \emph{remote reference}.

@@

We look at resources in distributed programming in more details in Section~\ref{sec:applications-flat-distr}.
The important point here is that in flat coeffect systems, contextual requirements are
\emph{split} between the call site and declaration site. Furthermore, there is no syntactic
structure that determines how the requirements are split. As mentioned in Section~\ref{sec:path-binding-amb},
we decouple the definition of semantics from the domain-specific choice that determines how
context demands are satisfied. We capture the choice in the type and give semantics over a
\emph{typing derivation}. A domain-specific algorithm then chooses the desirable typing -- for
example, by preferring resources available on the client over resources available on the server.

@@

In Section~\ref{sec:intro-why-array}, we used an example of tracking array access patterns. For every
variable, the additional coeffect annotation keeps a range of indices that may be accessed relatively
to the current cursor. For example, consider an expression
$x[\kvd{cursor}] = y[\kvd{cursor}-1] + y[\kvd{cursor}+1]$.

Here, the variable context $\Gamma$ contains two variables, both of type \ident{Arr}. This means
$\Gamma = x\!:\!\ident{Arr},\, y\!:\!\ident{Arr}$. For simplicity, we treat \kvd{cursor} as a
language primitive. The coeffect annotations will be $(0,0)$ for $x$ and $(-1,1)$ for $y$,
denoting that we access only the current value in $x$, but we need access to both left and right
neighbours in the $y$ array. In order to unify the flat and structural notions, we attach this information
as a \emph{vector} of annotations associated with a \emph{vector} of variable and write:
$\coctx{x\!:\!\ident{Arr},\, y\!:\!\ident{Arr}}{ \aclrd{\alift{ (0,0), (-1,1) }} }$.
The unification is outlined in Section~\ref{sec:further-unified}.

@@

Before looking at concrete flat and structural systems, we briefly overview
some notation used in this thesis. Structural coeffects keep annotations as \emph{vectors} and
use a number of operations related to scalars and vectors.

@@

Structural coeffects have one annotation per each variable. Thus, the annotation consists
of multiple values -- one belonging to each variable. To distinguish between the overall annotation
and individual (per-variable) annotations, we call the overall coeffect a \emph{vector} consisting of
\emph{scalar} coeffects. This asymmetry also explains why coeffects are not trivially dual to
effects.

@@

Just as in the usual pointwise multiplication of a vector by a scalar, we lift any binary operation on scalars into a
scalar-vector one. For a binary operation on scalars $\circ : \C \times \C \rightarrow \C$, we define
 $\cclrd{s} \circ \aclrd{\textbf{r}} = \alift{ \cclrd{s}\circ\cclrd{r_1}, \ldots, \cclrd{s}\circ\cclrd{r_n}}$.
Relations on scalars can be also lifted to vectors. Given two vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ of the
same shape with positions $\{ 1, \ldots, n \}$ and a relation $\varpropto\, \subseteq \C \times \C$ we define
$\aclrd{\textbf{r}} \varpropto \aclrd{\textbf{s}} \Leftrightarrow (\cclrd{r_1} \varpropto \cclrd{s_1}) \wedge \ldots \wedge (\cclrd{r_n} \varpropto \cclrd{s_n}) $
Finally, we often concatenate vectors, for example, when joining two variable contexts.
Given vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ with (possibly different) shapes $\{ 1, \ldots, n \}$ and
$\{ 1, \ldots, m \}$, the associative operation for concatenation $\vtimes$ is defined as
$\aclrd{\textbf{r}}\vtimes\aclrd{\textbf{s}} = \alift{\cclrd{r_1},\ldots,\cclrd{r_n},\cclrd{s_1},\ldots,\cclrd{s_m}}$.

@@

In flat coeffect systems, the additional contextual information are independent of lexically scoped
variables. As such, flat coeffects capture properties where the execution environment provides some
additional data, resources or information about the execution context.

@@

The fact that implicit parameters support both lexical and dynamic scoping becomes interesting
when we consider nested functions. The following function does some pre-processing and then returns a
function that builds a formatted string based on two implicit parameters $\ident{?width}$ and
$\ident{?size}$:

\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{format} = \lambda \ident{str}~\rightarrow \\[-0.25em]
\quad \kvd{let}~\ident{lines} = \ident{formatLines}~\ident{str}~\ident{?width}~\kvd{in}\\[-0.25em]
\quad (\lambda \ident{rest}~\rightarrow~\ident{append}~
         \ident{lines}~\ident{rest}~\ident{?width}~\ident{?size})
\end{array}
\end{equation*}
%
The body of the outer function accesses the parameter $\ident{?width}$, so it certainly requires a context
$\{ \ident{?width} \}$. The nested function (returned as a result) uses the parameter
$\ident{?width}$, but in addition also uses $\ident{?size}$. Where should the parameters used by the
nested function come from?

To keep examples in this chapter uniform, we do not use the Haskell notation and instead
write $\tau_1 \xrightarrow{r} \tau_2$ for a function that requires implicit parameters specified by $r$.
We also assume that are implicit parameters are of type $\ident{num}$, so the annotation can be a
simple set of names (rather than mapping from names to types). In a purely dynamically scoped system,
implicit parameters would have to be defined when the user invokes the nested function.
However, implicit parameters behave as a combination of lexical and dynamic scoping. This means
that the nested function can capture the value of $\ident{?width}$ and require just $\ident{?size}$.
The following shows the two options:
%
\begin{equation}
\tag{\emph{dynamic}}
\ident{string} \xrightarrow{ \{ \ident{?width} \} }
  (\ident{string} \xrightarrow{ \{ \ident{?width}, \ident{?size} \} } \ident{string})
\end{equation}
\vspace{-1em}
\begin{equation}
\tag{\emph{mixed}}
\ident{string} \xrightarrow{ \{ \ident{?width} \} }
  (\ident{string} \xrightarrow{ \{ \ident{?size} \} } \ident{string})
\end{equation}
%
This is not a complete list of possible typings, but it demonstrates the options. The (\emph{dynamic})
case requires the parameter \ident{?width} twice (the caller may provide different value, in which
case, the semantics needs to specify which value is preferred). In the (\emph{mixed}) case, the
nested function captures the \ident{?width} parameter available from the declaration site. Using
the latter typing, the function can be called as follows:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{formatHello} = (~\kvd{letdyn}~\ident{?width}=5~\kvd{in}~\ident{format}~\texttt{"Hello"})\\[-0.25em]
 \kvd{in}\,(~\kvd{letdyn}~\ident{?size} = 10~\kvd{in}~\ident{formatHello}~\texttt{"world"}~)
\end{array}
\end{equation*}
%
For different typings of \ident{format}, different ways of calling it are valid. This illustrates
the point made in Section~\ref{sec:applications-structure-lam} -- flat coeffect programs have
multiple typing derivations and the semantics depends on the domain-specific choice of preferred
typing. The following section shows how this looks in the type system for implicit parameters.

@@

Figure~\ref{fig:applications-flat-impl} shows a type system that tracks the set of expression's
implicit parameters. The type system uses judgements of the form $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$
meaning that an expression $e$ has a type $\tau$ in a free-variable context $\Gamma$ with a set
of implicit parameters specified by $\cclrd{r}$. The annotations $\cclrd{r},\cclrd{s},\cclrd{t}$ are
finite partial functions mapping implicit parameter names to types, \ie~$\cclrd{r},\cclrd{s},\cclrd{t} \subseteq
\ident{Names}$. The expressions include \ident{?param} to read implicit
parameter and \kvd{letdyn} to bind an implicit parameter. The types are standard, but functions are
annotated with the set of implicit parameters that must be available on the call site, \ie~
$\tau_1 \xrightarrow{\cclrd{s}} \tau_2$.

@@

In this section, we have seen how this affects both the type system and the semantics of the
language. In the type system, the (\emph{abs}) rule places context-requirements on both the
declaration site and the call site. For implicit parameters, this rule introduces non-determinism
in the type-inference, because the parameters can be split arbitrarily. However, as we show in the
next section, this is not always the case. Semantically, lambda abstraction \emph{merges} two
parts of context (implicit parameter dictionaries) that are provided by the call site and
declaration site.

@@

Distributed programming was used as one of the motivating examples for coeffects in
Chapter~\ref{ch:intro}. This section explores the use case. We look at rebindable resources and
cross-compilation. The structure of both is very similar to implicit parameters and type
class constraints, but they demonstrate that there is a broader use for coeffect systems.

@@

A practical language that supports rebindable resources is for example Acute \cite{app-distributed-acute}.
In the following example, we use the construct $\kvd{access}~\ident{Res}$ to represent
access to a rebindable resource named $\ident{Res}$. The following simple function accesses
a database together with a current date; then it filters from the database based on the date:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{recentNews} = \lambda () \rightarrow\\[-0.25em]
 \quad\kvd{let}~\ident{db} = \kvd{access}~\ident{News}~\kvd{in}\\[-0.25em]
 \quad\ident{query}~\ident{db}~\str{SELECT * WHERE Date > \%1}~(\kvd{access}~\ident{Clock})
\end{array}
\end{equation*}
%
When \ident{recentNews} is created on the server and sent to the client, a remote reference to
the database (available only on the server) must be captured. If the client device supports a
clock, then \ident{Clock} can be locally \emph{rebound}, \eg, to accommodate time-zone changes.
Otherwise, the date and time needs to be obtained from the server too.

~

The type system and semantics for rebindable resources are essentially the same as those for
implicit parameters. Primitive requirements are introduced by the $\kvd{access}$ keyword.
Lambda abstraction splits the requirements between declaration site (capturing remote reference)
and call site (representing rebinding). For this reason, we do not discuss the system in details
and instead look at other uses.

@@

\paragraph{Cross-compilation.}
A related issue with distributed programming is the need to target increasing number of diverse
platforms. Modern applications often need to run on multiple platforms (iOS, Android, Windows or
as JavaScript) or multiple versions of the same platform. Many programming languages are capable
of targeting multiple different platforms. For example, functional languages that can be compiled
to native code and JavaScript include, among others, F\#, Haskell and OCaml \cite{app-ocaml-js}.

Links \cite{app-distributed-links}, F\# WebTools and WebSharper \cite{app-fsharp-webapps,app-fsharp-webtools},
ML5 and QWeSST \cite{app-distributed-ml5, app-distributed-qwesst} and Hop \cite{app-hop-lang} go
further and allow including code for multiple distinct platforms in a single source file.
A single program is then automatically split and compiled to multiple target runtimes. This
posses additional challenges -- it is necessary to check where each part of the program can run
and statically guarantee that it will be possible to compile code to the required target
platform (safe \emph{multi-targetting}).

@@

Subcoeffecting and application are trivially dual to the rules for implicit parameters. We just track
supported environments using intersection as opposed to tracking required parameters using union.
However, this symmetry does not hold for lambda abstraction (\emph{abs}), which still uses \emph{union}.
This models the case when there are two ways of executing the function:
%
\begin{itemize}
\item The function is represented as executable code for an call site environment
  and is executed there, possibly after it is marshalled and transferred to another machine.
\vspace{-0.5em}
\item The function body is compiled for the declaration site environment; the value
  that is returned is a remote reference to the code and function calls are performed as remote invocations.
\end{itemize}
%
This example ignores important considerations -- for example, it is likely desirable to make this
difference explicit (\eg~using explicit wrapping of unevaluated expressions) and the implementation
also needs to be clarified. For a system that does this, see \eg~ML5 \cite{app-distributed-ml5}).
The key point of our brief example is that the algebraic structure of coeffect annotations may be more
complex and use, for example, $\cap$ for application and $\cup$ for abstraction.

@@

\subsection{Liveness analysis}
\label{sec:applications-flat-live}
Our next example shows the idea of coeffects from a different perspective. Rather than
keeping additional information independent of the variable context, we track properties about how
variables are used. Nevertheless, we still look at the left-hand side of $\vdash$ and the structure
of the typing rules and semantics will be very similar.

~

@@

\paragraph{Forward vs. backward \& may vs. must.}
Static analyses can be classified as either \emph{forward} or \emph{backward} (depending on how they
propagate information) and as either \emph{must} or \emph{may} (depending on what properties they
guarantee). Liveness is a \emph{backward} analysis -- the requirements are propagated from variables
to their declarations. The distinction between \emph{must} and \emph{may} is apparent when we look
at an example with conditionals:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{defaultArg}~= \lambda \ident{cond} \rightarrow \lambda \ident{input} \rightarrow\\[-0.25em]
\quad\kvd{if}~\ident{cond}~\kvd{then}~42~\kvd{else}~\ident{input}
\end{array}
\end{equation*}
%
Liveness analysis is a \emph{may} analysis meaning that it marks variable as live when it
\emph{may} be used and as dead if it is \emph{definitely} not used. This means that the variable
\ident{input} is \emph{live} in the example above. A \emph{must} analysis would mark the variable
only if it was used in both of the branches (this is sometimes called \emph{neededness} or
\emph{very busy} variable/expression).

@@

\paragraph{Type system.}
A type system that captures whole-context liveness annotates the context with value of a
two-point lattice $\mathcal{L} = \{ \cclrd{\ident{L}}, \cclrd{\ident{D}} \}$. The annotation \cclrd{\ident{L}} marks
the context as \emph{live} and \cclrd{\ident{D}} stands for a \emph{dead} context.
Figure~\ref{fig:applications-flat-livealg} (a) defines the ordering $\cclrd{\sqsubseteq}$, meet $\cclrd{\sqcup}$ and join
operations $\cclrd{\sqcap}$ of the lattice.

@@

\paragraph{Summary.}
Unlike with implicit parameters, lambda abstraction for liveness analysis does not introduce
non-determinism. It simply duplicates the context demands. However, this still matches the
property of coeffects that impurities cannot be delayed or thunked and attached just to the
function arrow -- we place requirements on both call site and declaration site.

@@

The Section~\ref{sec:intro-why-array} briefly demonstrated that we can treat array access as an
operation that accesses a context. In case of arrays, the context is neighbourhood of a current
location in the array specified by a cursor. In this section, we make the example more concrete,
using a simpler and better studied programming model, dataflow languages.

@@

@@

\paragraph{Semantics.}
The language discussed in this section is a \emph{causal} dataflow language. This means that
a computation can access \emph{past} values of the stream (but not future values). In the semantics,
we again need richer structure over the input.

@@

\paragraph{Summary.}
Type systems have been used in the context of dataflow languages, for example to check
initialization properties \cite{app-dataflow-init}, but to our knowledge, not for checking
the maximal number of required past values. Thus this section serves not just as an example,
but also shows how coeffects can lead to novel results.

@@

\subsection{Permissions and safe locking}
In the implicit parameters and dataflow examples, the context provides additional resources or
values that may be accessed at runtime. However, coeffects can also track \emph{permissions} or
\emph{capabilities} to perform some operation. We can invert the intuition behind liveness and
use it as a trivial example. When the context is live, it contains a \emph{permission} to access
variables. In this section, we briefly consider a system for safe locking of Flanagan and
Abadi~\cite{app-safe-locking} as one, more advanced example. Calculus of capabilities of
Cray et al.~\cite{app-capabilities} is discussed later in Section~\ref{sec:applications-active-ccc}.

>>>>

\paragraph{Chapter structure and contributions}
\begin{itemize}
\item We characterize contextual properties -- Section~\ref{sec:applications-structure}
  explains what is a \emph{coeffect} and contrasts it with a better known notion of
  \emph{effect}. It explains what is the nature of properties that can be tracked using
  coeffect systems presented in this thesis.

@@

\subsection{Effectful lambda abstraction}
\label{sec:applications-structure-lam}
As outlined in Section~\ref{sec:intro-theory}, the difference between effects and coeffects becomes 
apparent when we consider lambda abstraction. The typical lambda abstraction rule for effect systems 
looks as (\emph{effect}) in Figure~\ref{fig:applications-abs}. Wadler and 
Thiemann~\cite{monads-effects-marriage} explain how the effect analysis works as follows:
%
\begin{quote}
\emph{In the rule for abstraction, the effect is empty because evaluation immediately
returns the function, with no side effects. The effect on the function arrow
is the same as the effect for the function body, because applying the function will
have the same side effects as evaluating the~body.}
\end{quote}
%
This is the key property of \emph{output impurity}. The effects are only produced when the
function is evaluated and so the effects of the body are attached to the function. A recent
work by Tate~\cite{effects-producer-semantics} uses the term \emph{producer} effect systems
for such standard systems and characterises them as follows:
%
\begin{quote}
\emph{Indeed, we will define an effect as a producer effect if all computations with that
effect can be thunked as ``pure'' computations for a domain-specific notion of purity.}
\end{quote}
%
The thunking is typically performed by a lambda abstraction -- given an effectful expression
$e$, the function $\lambda x.e$ is an effect free value (thunk) that delays all effects.
As shown in the next section, contextual properties do not follow this pattern.

@@

The interesting case is when we construct a lambda function $\lambda x.e$, marshall it and
send it to another node. In systems such as Acute \cite{app-distributed-acute}, the context
requirements can be satisfied in a number of ways. When the same resource is available at the target
machine (\eg~current time), we can transfer the function with a context requirement and \emph{rebind}
the resource. However, if the resource is not available (\eg.~GPS on the server), we need to 
capture \emph{remote reference}.

@@

We look at resources in distributed programming in more detail in Section~\ref{sec:applications-flat-distr}.
The important point here is that in flat coeffect systems, contextual requirements are
\emph{split} between the call site and declaration site. Furthermore, there is no syntactic
structure that determines how the requirements are split. As mentioned in Section~\ref{sec:path-binding-amb},
we decouple the definition of semantics from the domain-specific choice that determines how
context demands are satisfied. We capture the choice in the type and give semantics over a
\emph{typing derivation}. A domain-specific algorithm then chooses the desirable typing -- for
example, by preferring resources available on the client over resources available on the server.

@@

In Section~\ref{sec:intro-why-array}, we used an example of tracking array access patterns. For every
variable, the additional coeffect annotation keeps a range of indices, relative to the current cursor,
that may be accessed. For example, consider an expression
$x[\kvd{cursor}] = y[\kvd{cursor}-1] + y[\kvd{cursor}+1]$.

Here, the variable context $\Gamma$ contains two variables, both of type \ident{Arr}. This means
$\Gamma = x\!:\!\ident{Arr},\, y\!:\!\ident{Arr}$. For simplicity, we treat \kvd{cursor} as a
language primitive. The coeffect annotations will be $(0,0)$ for $x$ and $(-1,1)$ for $y$,
denoting that we access only the current value in $x$, but we need access to both left and right
neighbours in the $y$ array. The context annotation is written as:
$\coctx{x\!:\!\ident{Arr},\, y\!:\!\ident{Arr}}{ \aclrd{\alift{ (0,0), (-1,1) }} }$.

Note that we attach the per-variable information as a \emph{vector} of annotations 
associated with a \emph{vector} of variables, which makes it possible to treat flat and structural
coeffects uniformly (Section~\ref{sec:struct-semantics} and \ref{sec:semantics-theory}) and unify
them into a single system (Section~\ref{sec:further-unified}), which is capable of tracking both
flat per-context coeffects and structural per-variable coeffects. Attaching structural coeffect
annotations directly to individual variables would simplify some aspects of our formalism; we choose
not to follow that approach as we envision that a fully unified theory of coeffects (which we 
attempt to approach in this thesis) would be able to capture variables, flat and structural coeffects 
using just a single mechanism.

@@

@@

Structural coeffects have one annotation per variable. Thus, the annotation consists
of multiple values -- one belonging to each variable. To distinguish between the overall annotation
and individual (per-variable) annotations, we call the overall coeffect a \emph{vector} consisting of
\emph{scalar} coeffects. This asymmetry also explains why coeffects are not trivially dual to
effects.

@@

Just as in the usual pointwise multiplication of a vector by a scalar, we lift any binary operation on scalars into a
scalar-vector one. For a binary operation on scalars $\circ : \C \times \C \rightarrow \C$, we define
 $\cclrd{s} \circ \aclrd{\textbf{r}} = \alift{ \cclrd{s}\circ\cclrd{r_1}, \ldots, \cclrd{s}\circ\cclrd{r_n}}$
 and $\aclrd{\textbf{r}}\circ\cclrd{s} = \alift{ \cclrd{r_1}\circ\cclrd{s}, \ldots, \cclrd{r_n}\circ\cclrd{s}}$.
Relations on scalars can be also lifted to vectors. Given two vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ of the
same shape with positions $\{ 1, \ldots, n \}$ and a relation $\varpropto\, \subseteq \C \times \C$ we define
$\aclrd{\textbf{r}} \varpropto \aclrd{\textbf{s}} \Leftrightarrow (\cclrd{r_1} \varpropto \cclrd{s_1}) \wedge \ldots \wedge (\cclrd{r_n} \varpropto \cclrd{s_n}) $
Finally, we often concatenate vectors, for example, when joining two variable contexts.
Given vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ with (possibly different) shapes $\{ 1, \ldots, n \}$ and
$\{ 1, \ldots, m \}$, the associative operation for concatenation $\vtimes$ is defined as
$\aclrd{\textbf{r}}\vtimes\aclrd{\textbf{s}} = \alift{\cclrd{r_1},\ldots,\cclrd{r_n},\cclrd{s_1},\ldots,\cclrd{s_m}}$.

@@

In flat coeffect systems, the additional contextual information is independent of lexically scoped
variables. As such, flat coeffects capture properties where the execution environment provides some
additional data, resources or information about the execution context.

@@

The fact that implicit parameters support both lexical and dynamic scoping becomes interesting
when we consider nested functions. The following function does some pre-processing and then returns a
function that builds a formatted string based on two implicit parameters $\ident{?width}$ and
$\ident{?size}$:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{format} = \lambda \ident{str}~\rightarrow \\[-0.25em]
\quad \kvd{let}~\ident{lines} = \ident{formatLines}~\ident{str}~\ident{?width}~\kvd{in}\\[-0.25em]
\quad (\lambda \ident{rest}~\rightarrow~\ident{append}~
         \ident{lines}~\ident{rest}~\ident{?width}~\ident{?size})
\end{array}
\end{equation*}
%
The body of the outer function accesses the parameter $\ident{?width}$, so it certainly requires a context
$\{ \ident{?width} \}$. The nested function (returned as a result) uses the parameter
$\ident{?width}$, but in addition also uses $\ident{?size}$. Where should the parameters used by the
nested function come from?

To keep examples in this chapter uniform, we do not use the Haskell notation and instead
write $\tau_1 \xrightarrow{r} \tau_2$ for a function that requires implicit parameters specified by $r$.
We also assume that implicit parameters are of type $\ident{num}$, so the annotation can be a
simple set of names (rather than mapping from names to types). In a purely dynamically scoped system,
implicit parameters would have to be defined when the user invokes the nested function.
However, implicit parameters behave as a combination of lexical and dynamic scoping. This means
that the nested function can capture the value of $\ident{?width}$ and require just $\ident{?size}$.
The following shows the two options:
%
\begin{equation}
\tag{\emph{dynamic}}
\ident{string} \xrightarrow{ \{ \ident{?width} \} }
  (\ident{string} \xrightarrow{ \{ \ident{?width}, \ident{?size} \} } \ident{string})
\end{equation}
\vspace{-1em}
\begin{equation}
\tag{\emph{mixed}}
\ident{string} \xrightarrow{ \{ \ident{?width} \} }
  (\ident{string} \xrightarrow{ \{ \ident{?size} \} } \ident{string})
\end{equation}
%
This is not a complete list of possible typings, but it demonstrates the options. The (\emph{dynamic})
case requires the parameter \ident{?width} twice (the caller may provide different value, in which
case, the semantics needs to specify which value is preferred). In the (\emph{mixed}) case, the
nested function captures the \ident{?width} parameter available from the declaration site. Using
the latter typing, the function can be called as follows:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{formatHello} = (~\kvd{letdyn}~\ident{?width}=5~\kvd{in}~\ident{format}~\texttt{"Hello"})\\[-0.25em]
 \kvd{in}\,(~\kvd{letdyn}~\ident{?size} = 10~\kvd{in}~\ident{formatHello}~\texttt{"world"}~)
\end{array}
\end{equation*}
%
For different typings of \ident{format}, different ways of calling it are valid. This illustrates
the point made in Section~\ref{sec:applications-structure-lam} -- flat coeffect programs have
multiple typing derivations and the semantics depends on the domain-specific choice of preferred
typing. The following section shows how this looks in the type system for implicit parameters.

@@

Figure~\ref{fig:applications-flat-impl} shows a type system that tracks the set of expression's
implicit parameters. The type system uses judgements of the form $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$
meaning that an expression $e$ has a type $\tau$ in a free-variable context $\Gamma$ with a set
of implicit parameters specified by $\cclrd{r}$. The annotations $\cclrd{r},\cclrd{s},\cclrd{t}$ are
sets of names, \ie~$\cclrd{r},\cclrd{s},\cclrd{t} \subseteq
\ident{Names}$. The expressions include \ident{?param} to read implicit
parameter and \kvd{letdyn} to bind an implicit parameter. The types are standard, but functions are
annotated with the set of implicit parameters that must be available on the call site, \ie~
$\tau_1 \xrightarrow{\cclrd{s}} \tau_2$.

@@

In this section, we have seen how this affects both the type system and the semantics of the
language. In the type system, the (\emph{abs}) rule places context-requirements on both the
declaration site and the call site. For implicit parameters, this rule means there the system does
not have the pincipal type property, because the parameters can be split arbitrarily. As we show in 
the next section, this is not always the case. Semantically, lambda abstraction \emph{merges} two
parts of context (implicit parameter dictionaries) that are provided by the call site and
declaration site.

@@

Distributed programming was used as one of the motivating examples for coeffects in
Chapter~\ref{ch:intro}. This section explores the use case. We look at rebindable resources and
cross-compilation. Both of these could be seen as an instance of implicit parameters, but we
present them separately to illustrate other forms that coeffect systems can have. The examples
given later (\eg~in Section~\ref{sec:applications-flat-dataflow}) show coeffects that cannot be
easily seen as implicit parameters.

@@

A practical language that supports rebindable resources is for example Acute \cite{app-distributed-acute}.
In the following example, we use the construct $\kvd{access}~\ident{Res}$ to represent
access to a rebindable resource named $\ident{Res}$. The following simple function accesses
a database together with a current date; then it filters from the database based on the date:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{localNews} = \lambda () \rightarrow\\[-0.25em]
 \quad\kvd{let}~\ident{db} = \kvd{access}~\ident{News}~\kvd{in}\\[-0.25em]
 \quad\ident{query}~\ident{db}~\str{SELECT * WHERE Location = \%1}~(\kvd{access}~\ident{GPS})
\end{array}
\end{equation*}
%
When \ident{localNews} is created on the server and sent to the client, a remote reference to
the database (available only on the server) must be captured. If the client device supports a
GPS, then \ident{GPS} can be locally \emph{rebound} to provide news for the reader's area.
Otherwise, the default location needs to be obtained from the server.

The type system and semantics for rebindable resources are essentially the same as those for
implicit parameters. Primitive requirements are introduced by the $\kvd{access}$ keyword.
Lambda abstraction splits the requirements between declaration site (capturing remote reference)
and call site (representing rebinding). For this reason, we do not discuss the system in detail
and instead look at other uses.

@@

\paragraph{Cross-compilation.}
A related issue with distributed programming is the need to target an increasing number of diverse
platforms. Modern applications often need to run on multiple platforms (iOS, Android, Windows or
as JavaScript) or multiple versions of the same platform. Many programming languages are capable
of targeting multiple different platforms. For example, functional languages that can be compiled
to native code and JavaScript include, among others, F\#, Haskell and OCaml \cite{app-ocaml-js}.

Links \cite{app-distributed-links}, F\# WebTools and WebSharper \cite{app-fsharp-webapps,app-fsharp-webtools},
ML5 and QWeSST \cite{app-distributed-ml5, app-distributed-qwesst} and Hop \cite{app-hop-lang} go
further and allow including code for multiple distinct platforms in a single source file.
A single program is then automatically split and compiled to multiple target runtimes. This
poses additional challenges -- it is necessary to check where each part of the program can run
and statically guarantee that it will be possible to compile code to the required target
platform (safe \emph{multi-targetting}).

@@

Subcoeffecting and application are the same as the rules for implicit parameters. We just track
supported environments using intersection as opposed to tracking required parameters using union.
However, this symmetry does not hold for lambda abstraction (\emph{abs}), which still uses \emph{union}.
This models the case when there are two ways of executing the function:
%
\begin{itemize}
\item The function is represented as executable code for a call site environment
  and is executed there, possibly after it is marshalled and transferred to another machine.
\vspace{-0.5em}
\item The function body is compiled for the declaration site environment; the value
  that is returned is a remote reference to the code and function calls are performed as remote invocations.
\end{itemize}
%
This example ignores important considerations -- for example, it is likely desirable to make this
difference explicit (\eg~using explicit wrapping of unevaluated expressions) and the implementation
also needs to be clarified. For a system that does this, see \eg~ML5 \cite{app-distributed-ml5}).
The key point of our brief example is that the algebraic structure of coeffect annotations may be more
complex and use, for example, $\cap$ for application and $\cup$ for abstraction.

@@

\subsection{Liveness analysis}
\label{sec:applications-flat-live}
Our next example shows the idea of coeffects from a different perspective. Rather than
keeping additional information independent of the variable context, we track properties about how
variables are used. Nevertheless, we still look at the left-hand side of $\vdash$ and the structure
of the typing rules and the semantics will be very similar.

@@

\paragraph{Forward vs. backward \& may vs. must.}
Static analyses can be classified as either \emph{forward} or \emph{backward} (depending on how they
propagate information) and as either \emph{must} or \emph{may} (depending on what properties they
guarantee). Liveness is a \emph{backward} analysis -- the requirements are propagated from variable
uses to their declarations. The distinction between \emph{must} and \emph{may} is apparent when we look
at an example with conditionals:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{defaultArg}~= \lambda \ident{cond} \rightarrow \lambda \ident{input} \rightarrow\\[-0.25em]
\quad\kvd{if}~\ident{cond}~\kvd{then}~42~\kvd{else}~\ident{input}
\end{array}
\end{equation*}
%
Liveness analysis is a \emph{may} analysis meaning that it marks variable as live when it
\emph{may} be used and as dead if it is \emph{definitely} not used. This means that the variable
\ident{input} is \emph{live} in the example above. A \emph{must} analysis would mark the variable
only if it was used in both of the branches (this is sometimes called \emph{neededness} or
\emph{very busy} variable/expression).

@@

\paragraph{Type system.}
A type system that captures whole-context liveness annotates the context with value of a
two-point lattice $\mathcal{L} = \{ \cclrd{\ident{L}}, \cclrd{\ident{D}} \}$ where 
$\cclrd{\ident{L}} \cclrd{\sqsubseteq} \cclrd{\ident{D}}$ (Figure~\ref{fig:applications-flat-livealg} (a)). 
The annotation \cclrd{\ident{L}} marks the context as \emph{live} and \cclrd{\ident{D}} stands 
for a \emph{dead} context.

@@

\paragraph{Summary.}
Unlike with implicit parameters, lambda abstraction for liveness analysis has the principal types
property. It simply duplicates the context demands. However, this still matches the
property of coeffects that impurities cannot be delayed or thunked and attached just to the
function arrow -- we place requirements on both call site and declaration site.

@@

Section~\ref{sec:intro-why-array} briefly demonstrated that we can treat array access as an
operation that accesses a context. In case of arrays, the context is the neighbourhood of a current
location in the array specified by a cursor. In this section, we make the example more concrete,
using a simpler and better studied programming model, dataflow languages.

@@

\paragraph{Algebra of annotations.}
The coeffect annotations used in the liveness example in Section~\ref{sec:applications-flat-live}
form a two-point lattice. The annotations for dataflow do not form a lattice (as the absorption
laws do not hold for addition and maximum), but the structure resembles that of \emph{semiring}.
As discussed later in Section~\ref{sec:flat-calculus-algebra}, the coeffect annotations for many 
of our examples form a semiring, but this is not the case in general.

The coeffect annotations for dataflow forms a tropical semiring with $\cclrd{+}$ as the multiplication,
$\cclrd{\textnormal{max}}$ as the addition, $\cclrd{-\infty}$ as the zero element and $\cclrd{0}$ as
unit. In this section, we do not use the $\cclrd{-\infty}$ element, but this can be used to annotate
unused variables as discussed in Section~\ref{sec:flat-calculus-examples}.

@@

\paragraph{Semantics.}
The language discussed in this section is a \emph{causal} dataflow language. This means that
a computation can access \emph{past} values of the stream but not future values. In the semantics,
we again need richer structure over the input.

@@

\paragraph{Summary.}
Type systems have been used in the context of dataflow languages to check
initialization properties \cite{app-dataflow-init} and resource-aware systems are capable
of tracking how required values are accessed in dataflow programming \cite{app-pixie}.
This section formulates the problem in terms of coeffects.

@@

\subsection{Permissions and safe locking}
In the implicit parameters and dataflow examples, the context provides additional resources or
values that may be accessed at runtime. However, coeffects can also track \emph{permissions} or
\emph{capabilities} to perform some operation. We can invert the intuition behind liveness and
use it as a trivial example. When the context is live, it contains a \emph{permission} to access
variables. In this section, we briefly consider a system for safe locking of Flanagan and
Abadi~\cite{app-safe-locking} as one, more advanced example. The calculus of capabilities of
Cray et al.~\cite{app-capabilities} is discussed later in Section~\ref{sec:applications-active-ccc}.
